---
title: "Cyclistic Analysis"
author: "Mike Quach"
date: "`r Sys.Date()`"
output: html_document
---
## Introduction
Cyclistic is a fictional bike-sharing company based in Chicago. The marketing director believes that the key to the company's future success is to boost the conversion of casual riders into annual members. In order to achieve this we must determine and understand how subscribers and casual riders use the service differently. 

## Ask 
The marketing analytics team will be operating under the guidance of three questions: 
<ol>
<li>How do annual members and casual riders use Cyclistic bikes differently?</li>
<li>Why would casual riders by Cyclistic annual membership?</li>
<li>How can Cyclistic use digital media to influence casual riders to become members?<li>
</ol>

The director has assigned us with the question: "How do annual members and casual riders use Cyclistic bikes differently?

## Hypothesis 
Going into the analysis I had an assumption that casual riders just did not ride as often or as long as annual members. I anticipated that casual riders only needed a bike rental for quick trips and did not use it often or long enough to justify the price tag of an annual subscription. 

## Findings 
Through our analysis we find that although casual riders made less trips with our bikes, they did on average ride longer than annual paid members by over 400%. We also see a spike in ride length for casual riders on Wednesdays. We also see that casual users and annual members have have different primary stations. 

### Tableau
I exported the cleaned data to a CSV from R and imported it into Tableau. This allowed me to come to some other findings through connections with visualizations. 

[Tableau Dashboard](https://public.tableau.com/views/CyclisticUsertypeComparisson/ComparisonofUsers?:language=en-US&publish=yes&:sid=&:redirect=auth&:display_count=n&:origin=viz_share_link)

We see from these visualizations that casual riders and annual members have different primary stations. We also see that in the summer months, interestingly enough we get an increase in unique trips starting in June and then sharply falling off after August. However, even with this spike we see that there is still nearly more than 2X the amount of annual member rides during August compared to casual riders. 

## Recommendations
Our findings tell us that our casual riders and our annual members see different value propositions in our product. 
Annual members value them for short rides and our casual riders prefer longer rentals. I assume these preferences are due to the price structure of the annual membership and the current cost per rental for casual riders. Annual members can ride whenever and therefore have the leisure of taking short rides for small errands where as casual riders will probably want to get the most out of the cost of the rental and only rent it when necessary for long rides. The marketing team can target these casual riders by emphasizing the cost savings of a membership yearly in comparison to X amount of daily bike rentals. This will help the casual rider justify the cost and also possibly find the value of being able to rent a bike for small errands. 

## R Walk Through Below 

```{r load tidyverse}
library(tidyverse)  

library(conflicted)
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

```{r import your datasets }
#Upload your datasets (csv files) here
q1_2020 <- read_csv("bike_data_2020_q1.csv")
q4_2019 <- read_csv("bike_data_2019_q4.csv")
q3_2019 <- read_csv("bike_data_2019_q3.csv")
q2_2019 <- read_csv("bike_data_2019_q2.csv")
```

```{r rename columns}
(q1_2020 <- rename(q1_2020
                   ,trip_id = ride_id
                   ,start_time = started_at
                   ,end_time = ended_at
                   ,from_station_name = start_station_name
                   ,from_station_id = start_station_id
                   ,to_station_name = end_station_name
                   ,to_station_id = end_station_id
                   ,usertype = member_casual

))
(q2_2019 <- rename(q2_2019
                   ,trip_id = `01 - Rental Details Rental ID`
                   ,start_time = `01 - Rental Details Local Start Time`
                   ,end_time = `01 - Rental Details Local End Time`
                   ,from_station_name = `03 - Rental Start Station Name`
                   ,from_station_id = `03 - Rental Start Station ID`
                   ,to_station_name = `02 - Rental End Station Name`
                   ,to_station_id = `02 - Rental End Station ID`
                   ,usertype = `User Type`
                   ,bikeid = `01 - Rental Details Bike ID`
                   ,gender = `Member Gender`
                   ,birthyear = `05 - Member Details Member Birthday Year`
                   ,tripduration = `01 - Rental Details Duration In Seconds Uncapped`
))
```

``` {r convert ride_id and rideable_type to character so that they can stack correctly}
q1_2020 <-  mutate(q1_2020, trip_id = as.character(trip_id)) 
q4_2019 <-  mutate(q4_2019, trip_id = as.character(trip_id)) 
q3_2019 <-  mutate(q3_2019, trip_id = as.character(trip_id)) 
q2_2019 <-  mutate(q2_2019, trip_id = as.character(trip_id)) 
```

```{r stack individual quarter data frames into one big data frame}
all_trips <- bind_rows(q1_2020, q4_2019, q3_2019, q2_2019)
head(all_trips)
```

``` {r remove outdated columns}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, tripduration, bikeid, rideable_type))
head(all_trips)
```

```{r reassign names to maintain consistency}
all_trips <-  all_trips %>% 
  mutate(usertype = recode(usertype
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```

```{r check observations}
table(all_trips$usertype)
```
```{r add additional date columns for aggregation}
all_trips$date <- as.Date(all_trips$start_time) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

```{r ride length calculation}
all_trips$ride_length <- difftime(all_trips$end_time,all_trips$start_time)
```

```{r convert ride_length to numeric}
# this allows you to run calculations on this 
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

```{r remove bad data}
all_trips_v2 <- all_trips[!(all_trips$from_station_name == "HQ QR" | all_trips$ride_length<0),]
```

```{r descriptive analysis. (all figures in seconds)}
summary(all_trips_v2$ride_length)
```

```{r compare members and casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = min)
```

```{r average ride time by each day for members vs casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
```


```{r fix the order of the days}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

```{r reordered average ride time by each day for members vs casual users  }
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
```

```{r ridership data by type and weekday}
all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(usertype, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(usertype, weekday)								# sorts
```

``` {r visualize the number of rides by rider type}
all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge")
```

``` {r visualization for average duration}
all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = usertype)) +
  geom_col(position = "dodge")
```

```{r export your data for further analysis}
# please be mindful of your working directory
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = 'avg_ride_length.csv')

write.csv(all_trips_v2, file = 'all_trips_v2.csv')
```
